---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document: 
    keep_md: true
---


## Loading and preprocessing the data

Download the file **activity.zip** from the repository into the current directory and unzip it.
```{r echo=TRUE, message=FALSE}
library(dplyr)
library(lattice)
df <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?

1. Calculating the total number of steps taken per day.
```{r echo=TRUE}
totalStepsByDate1 <- tapply(df$steps, df$date, FUN = sum, na.rm=TRUE)
```

2. Histogram of the total number of steps taken each day.
```{r echo=TRUE}
barplot(totalStepsByDate1, xlab="Date", ylab="Total number of steps", main="Total number of steps taken each day")
```

3. The mean and median of the total number of steps taken per day.
```{r echo=TRUE}
meanMedianByDate1 <- df %>% group_by(date) %>% summarize(mean = mean(steps, na.rm=TRUE), median = median(steps, na.rm=TRUE))
as.data.frame(meanMedianByDate1)
```

## What is the average daily activity pattern?

1. A time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).
```{r echo=TRUE}
avgStepsByInterval1 <- df %>% group_by(interval) %>% summarize(averageSteps = mean(steps, na.rm=TRUE))
with(avgStepsByInterval1,plot(strptime(sprintf("%04d", interval), format="%H%M"),averageSteps,type="l", xlab="Interval", ylab="Average number of steps taken", main="Average daily activity pattern"))
```

2. The 5-minute interval, on average across all the days in the dataset, which contains the maximum number of steps.
```{r echo=TRUE}
as.data.frame(avgStepsByInterval1[avgStepsByInterval1$averageSteps==max(avgStepsByInterval1$averageSteps),])[,1]
```

## Imputing missing values

1. Total number of missing values in the dataset.
```{r echo=TRUE}
nrow(df[is.na(df$steps),])
```

2. Strategy for filling in all of the missing values in the dataset.
- Averaging the steps by date we get some dates with no value.
```{r echo=TRUE}
as.data.frame(meanMedianByDate1[is.na(meanMedianByDate1$mean),])
```
- For instance let's look at steps taken at all interval on **2013-10-08**. By taking the unique from the output we can confirm that none of the intervals have steps.
```{r echo=TRUE}
unique(df[df$date=="2012-10-08","steps"])
```
- Averaging the steps by interval is a better option as we get all non-NA values.
```{r echo=TRUE}
unique(is.na(avgStepsByInterval1$averageSteps))
```
- Now let's take the average steps by each interval and keep it in a new variable **avgStepsByInterval2**. 
```{r}
avgStepsByInterval2 <- tapply(df$steps, df$interval, FUN = mean, na.rm=TRUE)
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in. Store the transformed data frame in a new variable **tdf**. Please note that average steps by each interval can be decimal values and hence they have been rounded.
```{r echo=TRUE}
tdf <- df
tdf$steps <- ifelse(is.na(tdf$steps),round(avgStepsByInterval2[as.character(tdf$interval)],digits=0),tdf$steps)
```

4. Histogram of the total number of steps taken each day.
```{r echo=TRUE}
totalStepsByDate2 <- tapply(tdf$steps, tdf$date, FUN = sum)
barplot(totalStepsByDate2, xlab="Date", ylab="Total number of steps", main="Histogram of the total number of steps taken each day")
```

- Mean and median total number of steps taken per day. These values now have all non-NA values and hence differ from the estimates from the first part of the assignment.
```{r echo=TRUE}
meanMedianByDate2 <- tdf %>% group_by(date) %>% summarize(mean = mean(steps, na.rm=TRUE), median = median(steps, na.rm=TRUE))
as.data.frame(meanMedianByDate2)
```
- The impact of imputing missing data on the estimates of the total daily number of steps is that now we have non-zero total number of steps each day which we earlier use to have.
```{r echo=TRUE}
NROW(totalStepsByDate2[totalStepsByDate2==0])
NROW(totalStepsByDate1[totalStepsByDate1==0])
```

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable **type** in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r echo=TRUE}
tdf$date <- as.Date(tdf$date) 
tdf$type <- as.factor(ifelse(weekdays(tdf$date) %in% c("Saturday","Sunday"),"weekend","weekday"))
```

2. Panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)
```{r echo=TRUE}
avgStepsByIntervalWeekdays <- tdf %>% group_by(interval,type) %>% summarize(averageSteps = mean(steps, na.rm=TRUE))
xyplot(avgStepsByIntervalWeekdays$averageSteps ~ avgStepsByIntervalWeekdays$interval | avgStepsByIntervalWeekdays$type, layout=c(1,2), type="l", xlab="Interval", ylab="Number of steps")
```
